One Model May Not Fit All: Subgroup Detection Using MOB

### 2023-10-10 _MOB_paper.Rnw
todos from meetin on okt 10

\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups \edc{of persons showing different} effects in a wide range of parametric models. It provides a versatile tool for detecting and explaining heterogeneity of intervention effects. In this tutorial paper, we provide an introduction to the general MOB framework. In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item response theory models. In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models. We apply GLMM trees to a dataset from a study of \citep{Demi09}, who compared longitudinal performance of siblings who did, and siblings who did not participate in the Head Start program. Using GLMM trees, we identify subgroups of families in which children show comparatively larger or smaller gains in performance following participation in Head Start. In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, \edc{i.e.~differential item functioning (DIF).
DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.} We show how a recently developed stopping criterion \citep{HennyDeba23} can be used to guide subgroup detection based on DIF effect sizes.\\}

@@

What you see displayed in Figures~\ref{fig:lmm_tree}, \ref{fig:math_tree}, and \ref{fig:read_tree}\todo{adjust if raschtree example changes (Mirka: maybe it would be helpful to have a simplified figure of a GLMERtree and a Raschtree early in the manuscript already, so we can refer to it here?)}, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure~\ref{fig:lmm_tree}) and for a binary Rasch model (Figures~\ref{fig:math_tree} and \ref{fig:read_tree}). These figures share the tree-structure in the upper part, but differ in their lower part. The starting point of the tree structure is at the very top of each tree. The very top of each tree is called the root node, so what you see are actually ``upside down'' trees with the root at the top and the branches at the 
bottom. The root node at the very top contains the entire sample of persons in the data set. From there, the persons are divided into subgroups.

@@

As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice. Together with a purification step \citep[see][for details]{HennyDeba23}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically. This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items. 
\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
For example, items that show DIF between different language groups can often be improved by means of making sure that in all translations the meaning is as similar as possible, that the words employed in the translations for the different languages are equally frequently used, etc. In other situations, sources of DIF might be harder to eliminate, so that often DIF items are excluded from a test. Either way, the measurement model needs to be refitted and its assumptions checked again after the final set of items has been decided upon and, in the case of modified items, administered again.

@@

@@

\todo[inline]{Write summary of findings.}

@@

not possible to detect certain patterns: XOR problem, known for all trees

@@

sample size dependence, similar ideas like MH trees also possible for other MOB?

>>>>

\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups \edc{of persons showing different} effects in a wide range of parametric models. It provides a versatile tool for detecting and explaining heterogeneity of intervention effects. In this tutorial paper, we provide an introduction to the general MOB framework. In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item response theory models. In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models. We apply GLMM trees to a dataset from a study of \citep{Demi09}\todo{take that out?}, who compared longitudinal performance of siblings who did, and siblings who did not participate in the Head Start program. Using GLMM trees, we identify subgroups of families in which children show comparatively larger or smaller gains in performance following participation in Head Start. In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, \edc{i.e.~differential item functioning (DIF).
DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.} We show how a recently developed stopping criterion \citep{HennyDeba23} can be used to guide subgroup detection based on DIF effect sizes.\\}

@@

What you see displayed in Figures~\ref{fig:lmm_tree}, \ref{fig:math_tree}, and \ref{fig:read_tree}\todo{adjust if raschtree example changes (Mirka: maybe it would be helpful to have a simplified figure of a GLMERtree and a Raschtree early in the manuscript already, so we can refer to it here?); Marjolein: please read and decide whether we do need a picture of a regression tree and/or mob tree here, if yes, simulate but with abstract x1 x2, no story}, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure~\ref{fig:lmm_tree}) and for a binary Rasch model (Figures~\ref{fig:math_tree} and \ref{fig:read_tree}). These figures share the tree-structure in the upper part, but differ in their lower part. The starting point of the tree structure is at the very top of each tree. The very top of each tree is called the root node, so what you see are actually ``upside down'' trees with the root at the top and the branches at the 
bottom. The root node at the very top contains the entire sample of persons in the data set. From there, the persons are divided into subgroups.

@@

As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice. Together with a purification step \citep[see][for details]{HennyDeba23} \todo{Caro can add more text on achoring, incl. word anchoring, here to prepare example 2 - only shortly mention here and explain in more detail later in the example (will become more clear in example)}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically. This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items. 
\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
For example, items that show DIF between different language groups can often be improved by means of making sure that in all translations the meaning is as similar as possible, that the words employed in the translations for the different languages are equally frequently used, etc. In other situations, sources of DIF might be harder to eliminate, so that often DIF items are excluded from a test. Either way, the measurement model needs to be refitted and its assumptions checked again after the final set of items has been decided upon and, in the case of modified items, administered again.

@@

\todo[inline]{Marjolein add here comment that typically affect of age is typically strongest; take out next part, but comment on it in discussion: CV would be a technique to see how stable results are, cite something about CV in general}

@@

\todo[inline]{Write summary of findings. -- now think: rather not discuss substantive findings, only properties of the methods}

@@

@@

sample size dependence, similar ideas like MH trees also possible for other MOB? \todo[inline]{possibly Mirka add something about visual interpretation? because effect sizes in mixed effects models are not very clear}
